Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-01 text/ms.tex
Major edits to ms
After meeting with Nick, also reorder and incorporate
the supplementary figs

Black swans are statistically improbable events that nonetheless occur ---
often with profound implications. While extremes in the physical environment,
such as monsoons and heat waves, are widely studied and increasing in magnitude
and frequency, it remains unclear the extent to which ecological populations
buffer or suffer from extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in ecological
process noise. We apply our model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
they tend to be driven by climate and severe winters, cycles of parasites and
predators, and interactions between these elements. They are more frequently
detected for populations with longer time series and lower levels of process
noise. We find little evidence of intrinsic life-history covariates. \textit{TODO close
with modelling and policy implications. Something like:
The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.}

@@

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid marking the mass extinctions of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising results in field experiments may be far more common than we assume
\citep{doak2008}, and a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence has been limited. What appeared to be heavy tails
in breeding bird abundance \citep{keitt1998}, could also be explained by mixing
of species in the original dataset \citep{allen2001}. A subsequent large scale
analysis of the Global Population Dynamics Database (GPDD) found little
evidence for heavy-tail distributions in longer abundance time series
\citet{halley2002}.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value distribution has been
applied to environmental data \citep[e.g.][]{katz2005}. This distribution
describes the most extreme event per time interval (e.g.~heaviest rainfall per
year), but requires long time series that can be condensed into time intervals.
(2) Fitting a catastrophic mixture distribution in a state space model allows
one to quantify the probability that population events are extreme but is also
data intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis did not quantify the probability of black
swans on a continuous scale or allow for standard population dynamics.

Here we develop a framework for identifying ecological black swans in
population dynamic process noise --- the stochastic jumps from time step to
time step. Our framework quantifies both the magnitude and probability of
heavy-tailed dynamics, allows for a range of population dynamic models, can
incorporate observation uncertainty, and is applicable to commonly available
lengths of abundance time series. We apply our model to hundreds of populations
from around the world across four taxonomic classes to address three questions:
(1) how frequent and strong are black swans in population dynamics across
taxonomy, (2) what are some verified causes of black swans in ecological time
series, and (3) are there characteristics of time series or intrinsic
life-history characteristics associated with heavy-tailed dynamics. We find
a high probability of ecological black swans across a wide range of taxa but in
a limited number of populations. When they occur, extreme climate, predation,
parasites, and their interactions are common documented causes. We find little
evidence that intrinsic life-history characteristics are related to the
probability of black swans.

@@

To obtain estimates of the probability and magnitude of black swans, we fitted
population dynamic models to hundreds of abundance time series from around the
world. The key addition of our models, compared to typical population models,
was estimating the shape of the process noise tails. We then compared how these
estimates related to taxonomy, time series properties, and life-history
characteristics, and for a subset of populations, we documented the established
causes of ecological black swans.

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Material). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values. Our final dataset contained
\NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
with a median of \medianTimeSteps~time steps (range of
\minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

We extended the Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated samples than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics \citep[Supporting Material]{gelman2014}. For
$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We fit four alternative population models to test if they systematically
changed our conclusions. (1) Autocorrelation has been suggested as a reason for
increased observed variability of abundance time series through time, which
could create apparent heavy tails \citep{inchausti2002}; therefore, we fit
a model that modelled the serial correlation in the residuals. (2) Previous
work has modelled abundance or growth rates without accounting for density
dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in
which we assumed density independence. (3) Observation error could bias
parameter estimates \citep{knape2012} or mask our ability to detect heavy tails
\citep{ward2007}; therefore, we fit a model where we allowed for a fixed
quantity of observation error ($0.2$ standard deviations on a log scale). (4)
The Gompertz model assumes that population growth rate declines linearly with
log abundance. We also fit an alternative model, the Ricker-logistic model,
which assumes that population growth rate declines linearly with abundance
itself (Supporting Material).

@@

Finally, we investigated a sample of populations that our method categorized as
having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
possible, we found the documented causes of ecological black swans in the
primary data source cited in the GPDD or in other literature describing the
population. These populations were chosen haphazardly with the purpose of
generating hypotheses about the causes of population dynamic black swans.

@@

We found strong but rare evidence for black-swan population dynamics. Defining
black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our
main Gompertz model found evidence for heavy tails most frequently for birds
(\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and
fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
were taxonomically widespread, occurring for at least one population in
\POrdersHeavy\% of the taxonomic orders recorded. Accounting for time series
length and partially pooling inference across taxonomic class and order with
a multilevel model, there was stronger evidence for black swans in insect
populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders
with the highest median probability of heavy tails were insect orders ---
however, there was considerable uncertainty in these estimates
(Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $\ge 10$ in
$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of
$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
also observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
in a population with 60 time steps than in one with 30 time steps
(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
increased time series length was small ($0.\pHeavyNSixty$ vs.\
$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the United Kingdom was thought to reduce population
productivity and make the population vulnerable when a red tide hit in 1968
\citep{potts1980}. The population experienced a black swan downswing, but the
new availability of nest sites resulted in a rapid population upswing
\citep{potts1980}. As another example, an interaction between environmental-
and predation-mediated population cycles are thought to have caused a downward
black swan for a water vole population \citep{saucy1994}. Other black swans
were the result of a sequence of extreme climate events on their own. For
example, severe winters in 1929, 1940--1942, and 1962--1963 were associated
with black swan downswings in grey heron abundance in the United Kingdom
\citep{stafford1971}. Our analysis finds that the last event was a combination
of two black swan events in a row and it took the population three times longer
to recover than predicted \citep{stafford1971}.

@@

We found strong evidence that the process noise in ecological time series was
heavy-tailed under  Gompertz or Ricker-logistic models in
\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
swans more frequently for longer time series and for populations with a smaller
magnitude of process error, as we might expect. However, we failed to find
strong evidence that black swans were associated with density dependence,
population growth rate, or lifespan. In cases that we verified, black swans
were often a result of the interaction between elements of extreme climate,
predators and parasite cycles, and strong changes in human pressures. Our
empirical results, sensitivity analyses, and simulation tests suggest that
estimating the tail shape of process noise is a viable method of detecting
black swan population dynamics and if anything will underestimate the
probability of black swans. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
\citep{easterling2000,ipcc2012}.

@@

We might expect to observe black swan dynamics in ecological time series
because of either unmodelled intrinsic properties of populations or extrinsic
forces acting on populations. Since a t-distribution can be formed by a mixture
of normal distributions \citep{gelman2014}, we could observe heavy tails if we
miss some underlying mixture of intrinsic processes \citep{allen2001}. That
process might be an aggregation of populations across space, or population
diversity, or some intrinsic change in population variability through time.
Extrinsic forces could also cause black swan dynamics
\citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
example, extreme climate, predation from or competition with other species
experiencing black swans, or sharp changes in human pressure such hunting,
fishing, or habitat destruction might cause black swans. Alternatively, the
interaction of multiple ``normal'' extrinsic forces could give rise to black
swan ecological dynamics. This could occur if the interaction nature is
synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
experience a rare alignment \citep{denny2009}.

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value more than five
standard deviations below the mean is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero. (c--e) Simulated population dynamics from
a Gompertz model with process noise drawn from t distributions with different
values of $\nu$. Coloured dots in panels c and d represent jumps with less than
a 1 in 1000 chance of occurring in a normal distribution. (f--h) Estimates of
$\nu$ from models fit to the times series in panels c--e. Shown are the
posterior samples (histograms), median and interquartile range of the posterior
(IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed
lines). Colour shading behind panels f--h illustrates the region of heavy
tails.}
\label{fig:didactic}
\end{center}
\end{figure}

>>>>

Black swans are statistically improbable events that nonetheless occur ---
often with profound consequences. While extremes in the physical environment,
such as monsoons and heat waves, are widely studied and increasing in magnitude
and frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in ecological
process noise. We apply our model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). We find little
evidence of intrinsic life-history covariates, but when they occur, they tend
to be extrinsically driven by climate and severe winters, cycles of parasites
and predators, and interactions between these elements. Black swans were more
frequently detected for populations with longer time series and lower levels of
process noise. The mere presence of black swans in population dynamics
highlights the importance of developing robust conservation and management
strategies --- particularly as the frequency and magnitude of climatic extremes
increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid marking the mass extinctions of the K-T
boundary TODO-REF \citep{harnik2012}. Today, climate extremes, in concert with
shifts in mean temperature, are expected to cause the greatest ecological and
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are are
frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent
to which ecological systems buffer or suffer from black swans
\citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. A survey of ecologists indicated that
surprising outcomes of field experiments are far more common than we assume
\citep{doak2008}, and a number of major environmental events on geologic and
contemporary time scales could be considered to have caused ecological black
swans \citep{nunez2012}.
% TODO: pull anecdote from that study
Black swans, or heavy tailed distributions, have been discovered in time-series
of marine microbe populations \citet{segura2013} and time to extinction for
experimental waterflea populations \citet{drake2014}. As far as we can tell
there have been two systematic surveys for ecological black swans, one in North
American breeding birds and the Global Population Dynamics Database, that found
little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
Nevertheless there are methodological challenges to their detection
\citep{allen2001,ward2007}.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist in higher taxa. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
heavy-tailed dynamics might exist, but our ability to detect them requires
further development of statistical tools. For example: (1) The generalized
extreme value distribution has been applied to environmental data
\citep[e.g.][]{katz2005}. This distribution describes the most extreme event
per time interval (e.g.~heaviest rainfall per year), but requires time series
that are sufficiently long to be condensed into time intervals. (2) Fitting
a catastrophic mixture distribution in a state space model allows one to
quantify the probability that population events are extreme but is also data
intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have been
competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis did not quantify the probability of black
swans on a continuous scale or allow for standard population dynamics.

Here we develop a framework for identifying ecological black swans in
population dynamic process noise --- the stochastic jumps from time step to
time step. Our framework quantifies both the magnitude and probability of
heavy-tailed dynamics, allows for a range of population dynamic models, can
incorporate observation uncertainty, and is applicable to commonly available
lengths of abundance time series. We apply our model to hundreds of populations
from across four taxonomic classes to address three questions: (1) how frequent
and strong are black swans in population dynamics across taxonomy, (2) what are
some verified causes of black swans in ecological time series, and (3) are
there characteristics of time series or intrinsic life-history characteristics
associated with heavy-tailed dynamics. We find a high probability of ecological
black swans across a wide range of taxa but in a limited number of populations.
When they occur, extreme climate, predation, parasites, and their interactions
are common documented causes. We find little evidence that intrinsic
life-history characteristics are related to the probability of black swans.

@@

To obtain estimates of the probability and magnitude of black swans, we fitted
population dynamic models to abundance time series from around the world. The
key novelty of our approach, compared to typical population models, was
estimating the shape of the process noise tails by measuring the degrees of
freedom ($\nu$) of a Student t-distribution.

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Material). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
Our final dataset contained \NPops~populations across \NOrders~taxonomic orders
and seven taxonomic classes with a median of \medianTimeSteps~time steps (range
of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t
= 1$. The model is density independent if $b = 1$, maximally density dependent
if $b = 0$, and inversely density dependent if $b < 0$. The process noise
$\epsilon_t$ is usually modelled as normally distributed with standard
deviation $\sigma$. In previous analyses of the GPDD, the Gompertz has proved
to be the most frequently parsimonious population model \citep{brook2006}.

We extended the Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution (Fig.~\ref{fig:didactic}a, b). For example, at $\nu
= 2$, the probability of drawing a value more than five standard deviations
below the mean is $1.8$\%, whereas the probability of drawing such a value from
a normal distribution is nearly zero. As $\nu$ approaches infinity the
distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
By estimating the value of $\nu$ we can quantity how heavy-tailed the process
noise deviations are.

% TODO: cite panels c, d in Fig 1

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated samples than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics 
(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Material). 
For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. (1) Autocorrelation has been suggested
as a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002}; therefore,
we fit a model that modelled the serial correlation in the residuals. (2)
Previous work has modelled abundance or growth rates without accounting for
density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler
model in which we assumed density independence. (3) Observation error could
bias parameter estimates \citep{knape2012} or mask our ability to detect heavy
tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed
quantity of observation error ($0.2$ standard deviations on a log scale). (4)
The Gompertz model assumes that population growth rate declines linearly with
log abundance. We also fit an alternative model, the Ricker-logistic model,
which assumes that population growth rate declines linearly with abundance
itself (Supporting Material).

@@

Finally, we investigated a sample of populations that our method categorized as
having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
possible, we found the documented causes of ecological black swans in the
primary data source cited in the GPDD or in other literature describing the
population. These exemplar populations were chosen haphazardly with the purpose
of generating hypotheses about the causes of population dynamic black swans.

@@

We found strong, but rare, evidence for black-swan population dynamics. By
defining black-swan dynamics as a greater than $0.5$ probability that $\nu
< 10$, our main Gompertz model found evidence for heavy tails most frequently
for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects
(\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs},
\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring in
\POrdersHeavy\% of taxonomic orders. Accounting for time series length and
partially pooling inference across taxonomic class and order with a multilevel
model, there was stronger evidence for black swans in insect populations than
is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the
highest median probability of heavy tails were insect orders --- however, there
was considerable uncertainty in these estimates
(Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Three of the four
population dynamic phenomena were unimportant:
% TODO: I don't like "unimportant"
including an autocorrelation structure in the residuals, modelling population
growth rates without density dependence, or modelling the population dynamics
as Ricker-logistic did not systematically alter our conclusions. However,
allowing for a fixed quantity of observation error decreased our estimates of
black-swan dynamics slightly, increasing the median estimate of $\nu$ from
$<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations,
although the majority of $\nu$ estimates remained qualitatively similar
(Fig.~\ref{fig:alt}).
Our simulation testing show that, if anything, our models underpredict the true
magnitude and probability of heavy tailed events --- especially given the
length of the time series in the GPDD (Figs~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
also observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
in a population with 60 time steps than in one with 30 time steps
(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
increased time series length was small ($0.\pHeavyNSixty$ vs.\
$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We assembled the documented causes of black swan population dynamics in seven
exemplar populations with high probabilities of heavy tails (Table 1). The
majority of documented events were downward black swans and the majority
involved a confluence of events. For example, a shortage of nest sites for the
shag long-necked bird in the United Kingdom was thought to reduce population
productivity and make the population vulnerable when a red tide hit in 1968
\citep{potts1980}. The population experienced a black swan downswing, but the
new availability of nest sites resulted in a rapid population upswing
\citep{potts1980}. As another example, an interaction between environmental-
and predation-mediated population cycles are thought to have caused a downward
black swan for a water vole population \citep{saucy1994}. Other black swans
were the result of a sequence of extreme climate events on their own. For
example, severe winters in 1929, 1940--1942, and 1962--1963 were associated
with black swan downswings in grey heron abundance in the United Kingdom
\citep{stafford1971}. Our analysis finds that the last event was a combination
of two black swan events in a row and it took the population three times longer
to recover than predicted \citep{stafford1971}.

@@

We found strong evidence that the process noise in ecological time series was
heavy-tailed under  Gompertz or Ricker-logistic models in
\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
swans more frequently for longer time series and for populations with a smaller
magnitude of process error, as we might expect. However, we failed to find
strong evidence that black swans were associated with the intrinsic ecological
phenomena of density dependence, population growth rate, or lifespan. In cases
that we verified, black swans were often a result of the interaction between
elements of extreme climate, predators and parasite cycles, and strong changes
in human pressures. Our empirical results, sensitivity analyses, and simulation
tests suggest that estimating the tail shape of process noise is a viable
method of detecting black swan population dynamics and if anything will
underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century \citep{easterling2000,ipcc2012}.

@@

We might expect to observe black swan dynamics in ecological time series
because of either unmodelled intrinsic properties of populations or extrinsic
forces acting on populations. Since a t-distribution can be formed by a mixture
of normal distributions \citep{gelman2014}, we could observe heavy tails if we
miss some underlying mixture of intrinsic processes \citep{allen2001}. That
process might be an aggregation of populations across space, or population
diversity, or some intrinsic change in population variability through time.
Extrinsic forces could also cause black swan dynamics
\citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
example, extreme climate, predation from (or competition with) other species
experiencing black swans, or sharp changes in human pressure such hunting,
fishing, or habitat destruction might cause black swans. Alternatively, the
interaction of multiple ``normal'' extrinsic forces could give rise to black
swan ecological dynamics. This could occur if the interaction is synergistic
\citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience
a rare alignment \citep{denny2009}.

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines that focus on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once
our ability to predict the spatial probability and covariates of ecological
black swans improves. Second, we can develop management policy that is robust
to heavy tails. For example, setting target population abundances that are
appropriately set back from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on ecological extremes
represents the next frontier, particularly in the face of increased climate
extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value more than five
standard deviations below the mean is 1.8\%, whereas the probability of drawing
such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
(c--e) Simulated population dynamics from a Gompertz model with process noise
drawn from t distributions with three different values of $\nu$. Coloured dots
in panels c and d represent jumps with less than a 1 in 1000 chance of
occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
to the times series in panels c--e. Shown are the posterior samples
(histograms), median and interquartile range of the posterior (IQR) (dots and
line segments), and the exponential prior on $\nu$ (dashed lines). Colour
shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}
